**Machine learning identifies unique taxa differentiating proximal and distal

### 2017-05-15 submission/manuscript.Rmd
manuscript discussion updates

* 2C: NMDS/ordination of some kind ??? maybe if i calculate inter/intrapersonal variability

@@

#### Random Forest classification models identify important OTUs on each side
To identify OTUs that are distinct at each biogeographical site, we constructed several Random Forest machine-learning models built on OTU abundances. The first model we built to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A). The constructed model used ((Xopt? need to follow up with this too)) features for the left and ((Xopt)) for the right. The models performed very well when classifying these samples (AUC X and Y). The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the right lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B). The model classifying the left lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4B). These results indicate that there are fine differences between the different sites of the colon, and that these can be traced down to specific OTUs on each side.

@@

### exit comparison models?

##draft of oncogenic OTU paragraph, this may go in results or discussion

Some of the healthy patients have Fusobacterium OTUs. The most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which BLASTs to _Fusobacterium nucleatum subspecies animalis_. It is the only OTU to align with a nucleatum strain- the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors. ((cite joe and niels work somewhere in here, maybe in the discussion?)). The _Fusobacterium_ positive samples were located on the right and left mucosa and ranged up to 1% of the sample. ((Table or something?)). OTU00152 is _Porphyromonas_ and the most frequent sequence in that OTU aligns to _Porphyromonas asacharolytica_, another bacterium commonly detected and isolated from colorectal tumors. OTU152 was only detected on the left mucosa, and in fact was one of the OTUs the classification model identified as separating left and right sides. Samples that were positive for _P. asacharolytica_ ranged in abundances from 0.01% - 16%. Finally, another commonly cancer-associated species, _Parvimonas micra_ was identified to be most similar to OTU248. Only 4 individuals had this bacteria present in their colons, three in the left mucosa and one in the right mucosa.

@@

Recap of results, interesting discussion points about diversity and previous studies of diversity along the gut axis of healthy people

Discussion of differentially abundant OTUs, how they fit the ecological regions of the gut (schematic?)
some of the very high OTUs BLAST to stool residents, like that one bacteriodes (OTU20)

Put this stuff into context with Sears PNAS paper

We detected _F. nucleatum_ and _P. asacharolytica_ in 7 and 5 of our patients, respectively. These bacteria have been shown to be predictive of colorectal cancer in humans (cite Niel). Interestingly, while _F. nucleatum_ was found on both sides of the colon, _P. asacharolytica_ was only detected in the left mucosa. Not much is known about the distribution of _P. asacharolytica_ but given its documented anaerobic characteristics, it might not be surprising that it resides in the less-oygen-rich left bowel (cite something). In studies examining bacteria on colorectal cancer tumors, _F. nucleatum_ is more commonly detected on right-sided tumors, and distribution of _F. nucleatum_ decreases along the colon to rectum (cite Mima 2016). Of the (8) individuals positive for _F. nucleatum_ in this study, the bacteria was spread across the right mucosa, left lumena and left mucosa. That said, in a previous study of 1,102 colorectal cancer cases, only 138 (11%) of them contained detectable _F. nucleatum_ at all (Mima 2016).

Fuso found on the right side? discussion about cancer-specific OTUS
cant pull out ETBF or pks+ ecoli with 16S
found fuso and porpyhromonas, but these are healthy people
what is rate of colonization with these bugs in healthy in literature?
of course presence does not precede cancer or IBD necessarily

>>>>

Similarly, when examining community structure between sites and sides, there is no clear pattern among individuals. Microbial communities do not separate on an ordination by the left or right sides in either the mucosal or lumenal samples (Figure 2C).

@@

#### Random Forest classification models identify important OTUs on each side
To identify OTUs that are distinct at each biogeographical site, we constructed several Random Forest machine-learning models built on OTU abundances. The first model we built to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A). The constructed model used ((Xopt? need to follow up with this too)) features for the left and ((Xopt)) for the right. The models performed very well when classifying these samples (AUC X and Y). The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the right lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B). The model classifying the left lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4C). These results indicate that there are fine differences between the different sites of the colon, and that these can be traced down to specific OTUs on each side.

@@

### Bacterial OTUs associated with cancer are found in healthy individuals 
Some of the healthy patients have Fusobacterium OTUs. The most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which BLASTs to _Fusobacterium nucleatum subspecies animalis_. It is the only OTU to align with a nucleatum strain- the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors. ((cite joe and niels work somewhere in here, maybe in the discussion?)). The _Fusobacterium_ positive samples were located on the right and left mucosa and ranged up to 1% of the sample. ((Table or something?)). OTU00152 is _Porphyromonas_ and the most frequent sequence in that OTU aligns to _Porphyromonas asacharolytica_, another bacterium commonly detected and isolated from colorectal tumors. OTU152 was only detected on the left mucosa, and in fact was one of the OTUs the classification model identified as separating left and right sides. Samples that were positive for _P. asacharolytica_ ranged in abundances from 0.01% - 16%. Finally, another commonly cancer-associated species, _Parvimonas micra_ was identified to be most similar to OTU248. Only 4 individuals had this bacteria present in their colons, three in the left mucosa and one in the right mucosa. Thus, potentially oncogenic OTUs can be found in our sample set of 20 healthy individuals.

@@

discussion of diversity results- different patterns in each person. ref previous studies of diversity along gut axis of healthy people. in this same paper ( I think) can also reference difference in NMDS... if I end up including that.

discussion of similarity results- may combine with previous paragraph here, prior studies, again underscore extreme variability of humans

Discussion of RF models, how they compare to FIT, Joe's check joes paper, how well does model do in L vs R for FIT?
when we look at models for CRC, cant predict R vs L, so create models

discuss host response papers? what does that mean.

Discussion of differentially abundant OTUs, how they fit the ecological regions of the gut (schematic?)
some of the very high OTUs BLAST to stool residents, like that one bacteriodes (OTU20) - not necessarily anything interesting there.

Put this stuff into context with Sears PNAS paper on sidedness

We detected _F. nucleatum_ and _P. asacharolytica_ in 7 and 5 of our patients, respectively. These bacteria have been shown to be predictive of colorectal cancer in humans (cite Niel) and have oncogenic properties in cell culture and in mice (cite). Interestingly, while _F. nucleatum_ was found on both sides of the colon, _P. asacharolytica_ was only detected in the left mucosa. Not much is known about the distribution of _P. asacharolytica_ but given its documented anaerobic characteristics, it might not be surprising that it resides in the less-oygen-rich left bowel (cite something). In studies examining bacteria on colorectal cancer tumors, _F. nucleatum_ is more commonly detected on right-sided tumors, and distribution of _F. nucleatum_ decreases along the colon to rectum (cite Mima 2016). Of the (8) (40%) individuals positive for _F. nucleatum_ in this study, the bacteria was spread across the right mucosa, left lumen and left mucosa. That said, in a previous study of 1,102 colorectal cancer cases, only 138 (11%) of them contained detectable _F. nucleatum_ at all (Mima 2016). In a separate study of patients with inflammatory bowel disease (IBD), 50% of 22 IBD patients and 17.6% of 34 people without disease contained _F. nucleatum_ in mucosal biopsies (Strauss 2011). These studies have the added caveat of being sampled from prepared or cleaned colons. Thus, the presence of _F. nucleatum_ in a healthy individual is not necessarily linked to the development of future colorectal cancers. Because of the spatial distribution of the _F. nucleatum_ in our sample set, we cannot develop a model for the role of _F. nucleatum_ in the healthy colon. Indeed, rather than the _F. nucleatum_ initiating tumorigenesis, these bacteria may instead arrive "late" to the tumor microenvironment and feed on tumors that have alread begun to develop (Flynn 2016).

what is rate of colonization with these bugs in healthy in literature?

discuss host response papers
